---
title: "Strengths"
author: "Sam Newman"
date: "2023-10-21"
output: html_document
---

Background:
The data comes from one qualitative question asked in a survey that was given to
a universities (named disclosed for confidentiality) 2nd year MPP students (n = 269) 
from 2019-2023. The question was free-response and asked students to list their strengths about the UCLA MPP 
program. This script preforms thematic analysis by tallying the total number 
of mentions a theme is mentioned. A bar chart is created showing the 5 most 
common themes and a table is produced displaying all the themes with their 
respective number of mentions. 

Outline: 
1. Clean Data
2. Analyze Data
3. Visualize Data
 
```{r}
library(tidyverse)
library(readxl)
library(writexl)
library(stringr)
library(gt)
library(webshot2)
```

```{r}
setwd("C:/Grad School/Student Affairs/Exit Survey")
```


# load clean exit survey data 2019-2023 (created in another script)
```{r}
sub_e2019_2023 <- readRDS("sub_e2019_2023.rds")
```


*Clean Variable*

# subset data and investigate subset 
```{r}
#create subset of Themes_Strengths
subset_strengths<- sub_e2019_2023 %>%  
  select(Themes_Strengths)

#269 observations 
nrow(subset_strengths)

#11 NA values 
sum(is.na(subset_strengths))

#reduces observations to 258 (practice)
subset_strengths %>%  
  select(Themes_Strengths) %>%  
  filter(!is.na(Themes_Strengths)) %>%  
  nrow()

#remove NAs
clean_subset_strengths <- subset_strengths %>%  
  select(Themes_Strengths) %>%  
  filter(!is.na(Themes_Strengths))

#observations are 258. Code worked. 
nrow(clean_subset_strengths)

#124 unqiue answers. Although this will change once I seperate entries by commas 
unique(clean_subset_strengths$Themes_Strengths)

#Class is character
class(clean_subset_strengths$Themes_Strengths)
```


# seperate values by commas
```{r}
comma_seperated_strengths <- clean_subset_strengths %>%  
  separate_rows(Themes_Strengths, 
                sep = ",")

#object now has 451 observations 
length(comma_seperated_strengths$Themes_Strengths)
```

# clean comma_seperated_strengths
```{r}
#remove white space before and after values 
clean_comma_seperated_strengths <- comma_seperated_strengths %>% 
  mutate(Themes_Strengths = 
           str_squish(Themes_Strengths))
 
#convert all letters to lower case 
clean_comma_seperated_strengths <- clean_comma_seperated_strengths %>%  
  mutate(Themes_Strengths = 
           tolower(Themes_Strengths))

#33 unique values 
unique(clean_comma_seperated_strengths$Themes_Strengths)

#recode misspelled values 
clean_comma_seperated_strengths <- clean_comma_seperated_strengths %>%
  mutate(Themes_Strengths = 
           recode(Themes_Strengths,
                  
                                   " " = 'NA',
                                   "alumni connections" = "career resources", 
                                   "alumni networking" = "career resources", 
                                   "career networking" = "career resources", 
                                   "career resourses" = "career resources", 
                                   "classes" = "courses", 
                                   "classses outside of luskin" = "courses",
                                   "core classses" = "core classes", 
                                   "na" = 'NA', 
                                   "quantiative" = "quantitative", 
                                   "quntitative" = "quantitative", 
                                   "realtionships" = "relationships", 
                                   "relationship" = "relationships",
                                   "department" = "relationships",
                                   "faculty" = "relationships",
                                   "leanring from professors" = "courses", 
                                   "practioners" = "practical skills", 
                                   "pratical skills" = "practical skills", 
                                   "applied learning" = "practical skills"))
              
#17 unique values 
unique(clean_comma_seperated_strengths$Themes_Strengths)

#check code to remove NAs 
clean_comma_seperated_strengths %>%  
  filter(Themes_Strengths != "NA") %>%
  unique()

#check code to remove blank values 
clean_comma_seperated_strengths %>%  
  filter(Themes_Strengths != "") %>% 
  unique()

#remove NAs and blank values 
clean_comma_seperated_strengths <-  clean_comma_seperated_strengths %>%  
  filter(Themes_Strengths != "NA", 
         Themes_Strengths != "",
         Themes_Strengths != " ")


#14 unique values 
unique(clean_comma_seperated_strengths$Themes_Strengths)
```

*Analyze variable* 

# add count and percent to the number of times a value is repeated 
```{r}
#create object containing count 
count_comma_seperated_strengths <- clean_comma_seperated_strengths %>% 
  group_by(Themes_Strengths) %>%  
  summarize(count = n())

#create object containing percent 
count_comma_seperated_strengths <- count_comma_seperated_strengths %>%  
  mutate(percent = 
           round(count / sum(count), 4))

#order objects by numerically 
count_comma_seperated_strengths <- count_comma_seperated_strengths %>%  arrange(-count)

#number of observations is 443
sum(count_comma_seperated_strengths$count)
```
# keep top 5 counts 
```{r}
top5_strenghts <- count_comma_seperated_strengths[1:5,]
```

# change to uppercase and change the names of the variables
```{r}
#Capitalize the first letter of each word 
top5_strenghts <- top5_strenghts %>%  
  mutate(Themes_Strengths = 
           str_to_title(Themes_Strengths))
                                           
#change the name of courses 
top5_strenghts <- top5_strenghts %>%  
  mutate(Themes_Strengths = 
           recode(Themes_Strengths,
                   "Courses" = "Electives", 
                   "Quantitative" = "Quantitative Skills"))
```

# factor bars (preparing for visualization) 
```{r}
factor_top5_strengths <- top5_strenghts

factor_top5_strengths$Themes_Strengths <- factor_top5_strengths$Themes_Strengths %>%  
  factor(
  factor_top5_strengths$Themes_Strengths, 
  levels = c(
             "Electives",
             "Relationships", 
             "Quantitative Skills", 
             "Core Classes", 
             "Career Resources"
                               ))

#check class type (it's factor - good)
class(factor_top5_strengths$Themes_Strengths)

```

*Visualize Data* 

# create horizontal bar chart
```{r}
ggplot(factor_top5_strengths, 
       aes(x = Themes_Strengths, 
           y = count)) +
  
  geom_bar(
         stat = 'identity',
         fill = "lightblue",
         position = 'dodge', #makes bar chart side by side
         color = "black", #adds outline color to bins
         na.rm = TRUE)+
  
  scale_y_continuous(
                    breaks = seq(0,100, 20)) +
  labs( 
        x = "", 
        y = "Number of Mentions",
        title = "Strengths of MPP Program \n (2019-2023)")+
  theme(
        axis.text = element_text(hjust = .5, size =10), 
        plot.caption = element_text(hjust = 1,), 
        plot.title = element_text(hjust = .5, face = "bold", size = 12))+
        labs(caption ="443 mentions from 258 students") 
```

# create table showing all strengths for count_sep_strengths 
```{r}
#create row for total count 
new_row <- data.frame(
  Themes_Strengths = "Total", 
  count = sum(count_comma_seperated_strengths$count),
  percent = NA)

#add row to dataframe 
table_full_strengths <- add_row(count_comma_seperated_strengths, new_row)
```

# clean table_strengths
```{r}
#Format names of Themes Strength and capitalized variables name
clean_table_full_strengths <- table_full_strengths %>%  
  mutate(
    Themes_Strengths = str_to_title(Themes_Strengths), 
    Themes_Strengths = recode(Themes_Strengths, 
                              "App" = "APP",
                              "Dei" = "DEI")) %>%  
  rename(
    Mentions = count, 
    Percent = percent, 
    Strengths = Themes_Strengths)

#remove percent 
clean_table_full_strengths <- clean_table_full_strengths %>%  select(-Percent)
```

# create table 
```{r}
strengths_table <- gt(clean_table_full_strengths) 
```

# create file directory and save table in directory as .png 
```{r}
#create object that points to folder where I want table to be saved
table_dir <- "C:/Grad School/Student Affairs/Exit Survey/Visualizations"

#save table to that directory 
gtsave(strengths_table, filename = file.path(table_dir, "strengths_table.png"))
```

