---
title: "fundingMpp_BestFinancialAid"
author: "Sam Newman"
date: "2023-09-30"
output: html_document
---

Background:
The data comes from one qualitative question asked in a survey that was given to
a universities (named disclosed for confidentiality) 2nd year MPP students (n = 269) 
from 2019-2023. The question asked to list all the funding they received at the 
university, and was a free response question.  This script cleans the data using 
the stringr package and creates a horizontal bar chart displaying the top 9 most 
mentioned types of funding. 

Outline: 
1. Clean Data
3. Visualize Data

```{r}
library(tidyverse)
library(readxl)
library(writexl)
library(stringr)
library(gt)
```

```{r}
setwd("C:/Grad School/Student Affairs/Exit Survey")
```
# load clean exit survey data 2019-2023 (created in another script)
```{r}
sub_e2019_2023 <- readRDS("sub_e2019_2023.rds")
sub_e2019_2023 %>%  select(best_course) %>% view()
```

*Clean Variable*

# create subset for data and intial investigation  
```{r}
funding_MPP <- sub_e2019_2023 %>%  select(funding_MPP, year)

#269 observations, class is character 
summary(funding_MPP)

#All the values are separated by commas. Some values that are free response. 
unique(funding_MPP$funding_MPP)

#rows that funding_MPP NAs: 157 160 162 175
which(is.na(funding_MPP$funding_MPP))

#no NAs
which(is.na(funding_MPP$year))

#confirming values are NA. They are all in year 2019
funding_MPP[c(157, 160, 162, 175),] %>%  view()

#remove values that NA for "funding_MPP" and "year" 
 rm_na <- funding_MPP %>%  
   filter(is.na(funding_MPP))

funding_clean <- funding_MPP %>%  
  filter(!is.na(funding_MPP))

#there are 0 NAs left. 
sum(is.na(funding_clean$funding_MPP))

#There are 265 observations. 
length(funding_clean$funding_MPP)

#view distribution of responses by year 
funding_clean %>%  
  group_by(year) %>%  
  summarize (count = n())
```
Clean data to seperate by commas using stringr

#  change "on and off campus" to "on campus and off campus"
```{r}
on_off_campus <- which(!is.na(str_match(
  funding_clean$funding_MPP, 
  "on and off campus"))
  )

funding_clean[on_off_campus, 1] %>%  view()

#check code to make sure it works 
str_replace_all(funding_clean$funding_MPP,
                "on and off campus", 
                "on campus employment and off campus") 

funding_clean_2 <- funding_clean %>% 
  mutate(funding_MPP = 
           str_replace_all(funding_MPP, 
                           "on and off campus", 
                           "on campus employment and off campus"))

#check values 
funding_clean_2$funding_MPP[c(36,84)]
```
# replace "and" with commas
```{r}
#which values use an "and"  
and_rows <- which(!is.na(str_match(funding_clean$funding_MPP, "and"))) 

#view and_rows
funding_clean[and_rows, 1] %>%  view()

#check code 
str_replace_all(funding_clean$funding_MPP,
                "and", 
                ",") 

#remove 'and' from object and replace with comma 
funding_clean <- funding_clean %>% 
  mutate(funding_MPP = 
           str_replace_all(funding_MPP,
                           "and",
                           ","))

#check values 
which(!is.na(str_match(funding_clean$funding_MPP, "and"))) 
```

# save object to use in FundingMPP_NumberOfSources 
```{r}
save(funding_clean, file = "fundingMPP_clean")
```

# analyize Theme_Improvements 

# seperate values by commas
```{r}
funding_sep <- funding_clean %>%  
  separate_rows(funding_MPP, 
                sep = "[,.]")


funding_sep %>%  
  group_by(funding_MPP) %>% 
  summarize (count = n()) 
```

# clean funding_sep
```{r}
#remove white space before and after comma separated values 
funding_sep_clean <- funding_sep %>%  
  mutate(funding_MPP =  
           str_squish(funding_MPP))

#convert all words to lower case 
funding_sep_clean <- funding_sep_clean %>%  
  mutate(funding_MPP =  
           tolower(funding_MPP))

#these rows have blank values (18 total)
which(funding_sep_clean$funding_MPP == "") 
length(which(funding_sep_clean$funding_MPP == ""))

#make sure code works to delete blank values  
funding_sep_clean %>%  
  filter(funding_MPP != "") %>%  
  nrow()

length(funding_sep_clean$funding_MPP)

#remove blank rows 
funding_sep_clean <- 
  funding_sep_clean %>%  
  filter(funding_MPP != "")

#check that values were removed
which(funding_sep_clean$funding_MPP == "") 
```

# make count object of funding_sep_clean
```{r}
funding_sep_count <- funding_sep_clean %>%  
  group_by(funding_MPP) %>%  
  summarize (count = n()) %>%  
  arrange(count) %>%  
  view()
```

# recode misspelled words 
```{r}
funding_sep_count_clean <- funding_sep_count %>%
  mutate(funding_MPP = recode(funding_MPP,
              "family resources" = "external sponsor (family/partner support)",
              "family resources." = "external sponsor (family/partner support)",
              "family" = "external sponsor (family/partner support)",
              "family resources for regular expenditures" = "external sponsor 
              (family/partner support)",
              "family supports." = "external sponsor (family/partner support)",
              "fellowships" = "department fellowship/scholarship",
              "fellowship" = "department fellowship/scholarship",
              "fellowships/awards" = "department fellowship/scholarship",
              "department fellowships both years" = "department 
              fellowship/scholarship",
              "graduate opportunity fellowship" = "department 
              fellowship/scholarship",
              "i received awards" = "department fellowship/scholarship",
              "graduate opportunity fellowship program for the first year" 
              = "department fellowship/scholarship",
              "awards" = "department fellowship/scholarship", 
              "scholarships" = "department fellowship/scholarship", 
              "loans" = "student loans",
              "loans." = "student loans",
              "most funding was from loans." = "student loans",
              "gsrs" = "gsr",
              "gsr position" = "gsr", 
              "gsr position both years" = "gsr",
              "a luskin fellowship" = "luskin fellowship/scholarship 
              (david bohnett fellowship/liberty hill fellowship/dukakis 
              fellowship)",
              "david bohnett fellowship" = "luskin fellowship/scholarship 
              (david bohnett fellowship/liberty hill fellowship/dukakis fellowship)",
              "fellowships from the luskin school" =  
                "luskin fellowship/scholarship (david bohnett 
              fellowship/liberty hill fellowship/dukakis fellowship)",
              "off campus employment" = "off-campus employment", 
              "interned for half of my first year"= "off-campus employment", 
              "later as a paid intern at a government agency for part of the 
              time that i was in the program" = "off-campus employment", 
              "internship" = "off-campus employment",
              "company" = "off-campus employment", 
              "covered by my employer" = "off-campus employment", 
              "on campus employment" = "on-campus employment",
              "a paid internship." = "off-campus employment", 
              "my time here was funded by the u.s. army" = "off-campus employment",
              "a pickering fellowship" = "external fellowship",
              "deans leaders in health" = "external fellowship",
              "md (prime-la) funding" = "external fellowship",
              "med school scholarship/loans are included" = "external fellowship",
              "i also received scholarships through the medical school" 
              = "external fellowship",
              "funded by government" = "external fellowship",
              "funding from the government" = "external fellowship",
              "funding from the government" = "external fellowship",
              "japan government" = "external fellowship",
              "japanese government fund" =  "external fellowship",
              "japanese government support" = "external fellowship",
              "gi bill" = "external fellowship",
              "government funding" = "external fellowship",
              "i ta'd for one quarter" = "taships", 
              "i was only given one taship" = "taships", 
              "csun professor" = "external fellowship",
              "alumni" = "department fellowship/scholarship",
              "inheritance" = "private saving",
              "my bank account" = "private saving",
              "i also worked part-time for asucla" = "off-campus employment", 
              "i typically had at least one job at all times" = "on-campus 
              employment", 
              "i was also employed as a reader at ucla" =  "on-campus
              employment",
              "i was only given one taship" = "taships", 
              "off campus internships" =  "off-campus employment", 
              "on campus" = "on campus employment", 
              "parent's support" = "external sponsor (family/partner support)",
              "parents" = "external sponsor (family/partner support)",
              "personal resources" = "private saving", 
              "personal savings" = "private saving", 
              "ppia fellowship" = "external fellowship",
              "private support" = "external sponsor (family/partner support)",
              "received a financial stipend" = "external fellowship",
              "received fellowships" = "department fellowship/scholarship",
              "received two fellowships from the department" = 
                "department      fellowship/scholarship",
              "savings from work experience" = "private saving", 
              "school of medicine prime program" = "external fellowship",
              "science scholarship"  = "external fellowship",
              "significant student loans!" = "student loans", 
              "simultaneous full-time employment - office of us senator alex 
              padilla" = "off-campus employment", 
              "sponsored by government" = "external fellowship",
              "sponsorship by my employer" = "off-campus employment", 
              "sponsorship from my country for tuition" =  "external fellowship",
              "sponsorship from the japanese government" = "external fellowship",
              "student loans mostly (almost $100" = "student loans",
              "taship" = "taships", 
              "the rest was loans" = "student loans", 
              "the ucla dean of students office" = "external fellowship",
              "the ucla graduate opportunity fellowship paid for my first year
              in graduate school" = "department fellowship/scholarship",
              "the very small fellowship from the department" = 
                "department fellowship/scholarship",
              "took out graduate student loans" = "student loans", 
              "ucla affiliates" = "department fellowship/scholarship",
              "ucla grad division" = "external fellowship",
              "worked as a graduate student researcher" = "gsr"
              ))
          
```

# recalculate count 
```{r}
funding_sep_count_clean2 <- funding_sep_count_clean %>%  
  group_by(funding_MPP) %>%  
  summarize (count = sum(count)) %>% 
  arrange(count) %>%  
  view()
```

# continue recoding values  
```{r}
funding_sep_count_clean2 <- funding_sep_count_clean2 %>%
  mutate(funding_MPP = recode(funding_MPP,
          "luskin fellowship/scholarship (david bohnett fellowship/liberty hill 
          fellowship/michael s" = "luskin fellowship/scholarship (david bohnett 
          fellowship/liberty hill fellowship/dukakis fellowship)",
          "luskin fellowship/scholarship (david bohnett fellowship/liberty 
          hill fellowship/michael s. dukakis fellowship)" = "luskin 
          fellowship/scholarship (david bohnett fellowship/liberty hill 
          fellowship/dukakis fellowship)",
          "campus employment other that ta" = "on-campus employment", 
          "gsr" = "on-campus employment", 
          "a paid internship" = "off-campus employment", 
          "army" = "off-campus employment",
          "family supports" = "external sponsor (family/partner support)",
          "luskin fellowships" = "luskin fellowship/scholarship (david bohnett 
          fellowship/liberty hill fellowship/dukakis fellowship)",
          "most funding was from loans" = "student loans", 
          "on campus employment" = "on-campus employment", 
          "raships" = "on-campus employment", 
          "scholarship"= "department fellowship/scholarship",
          "significant student loans! i also worked off" = "student loans", 
          "employment" = "off-campus employment",
          "private saving" = "private savings", 
          "taships" = "on-campus employment"
          ))
```

# recalculate count 
```{r}
funding_sep_count_clean3 <- funding_sep_count_clean2 %>%  
  group_by(funding_MPP) %>%  
  summarize (count = sum(count)) %>% 
  arrange(count) %>% 
  view()
```

# capitalize first letters of words  
```{r}
#check code  
funding_sep_count_clean3 %>%  
  mutate(funding_MPP = 
           str_to_title(funding_MPP))

clean_funding_sep_count_clean3 <- funding_sep_count_clean3 %>%  
  mutate(funding_MPP = 
           str_to_title(funding_MPP)) %>%  
  arrange(count) %>%  
  view()
```

# add percentage to count 
```{r}
clean_funding_sep_percent <- funding_sep_count_clean3 %>%  
  mutate(percent = 
           count/sum(count)) %>% 
  arrange(desc(count)) %>%  
  view()
```

*Data Visualization*

#factor bars 
```{r}
clean_funding_sep_percent$funding_MPP <- 
  factor(clean_funding_sep_percent$funding_MPP,
         levels = c(
           "External Fellowship",
           "Family/Partner Support",
           "Private Savings",
           "Student Loans",
           "Off-Campus Employment",
           "On-Campus Employment",
           "Department Fellowship/Scholarship"
           ))
```



# create grouped bar chart for all years
```{r}
ggplot(clean_funding_sep_percent, 
       aes(x = funding_MPP, 
           y = count)) +
                                      
  geom_bar(
    stat = 'identity',
    fill = "lightblue",
    position = 'dodge',
    color = "black", 
    na.rm = TRUE) +
  
  coord_flip() + 
  
  scale_y_continuous(
    limits = c(0, 250),                
    breaks = seq(0,250, 50)) +
  
  labs(
    x = "", 
    y = "Number of Mentions",
    title = "Types of Funding MPP Students Receive\n (2019-2023)") +
  
  theme(
    axis.text = element_text(hjust = .5, size =9), 
    plot.caption = element_text(hjust = 1, ), 
    plot.title = element_text(hjust = .5, face = "bold", size = 12)) +
  
  labs(
    caption ="820 mentions from 265 students") 
 
```

