---
title: "Funding_Debt_BeforeAfter"
author: "Sam Newman"
date: "2023-09-29"
output: html_document
---

Background:
The data comes from two quantitative questions asked in a survey that was given to
a universities (named disclosed for confidentiality) 2nd year MPP students (n = 269) 
from 2019-2023. Students checked a box stating their debt before and after the MPP 
program from one of the following boxes: 0, Less than $10,000, $10,000-$20,000, 
$30,000-$40,000, $40,000-$50,000, $50,000-$60,000, $60,000-$70,000, $70,000-$80,000, 
$80,000-$90,000, $90,000-$100,000, More than $100,000). This script consolidates 
the grouping of debt values, displays the changes in a grouped bar chart, and 
produces descriptive statistics.   


Outline: 
1. Clean Data 
2. Analyze Data 
3. Manipulate Data
4. Visualize Data
5. Descriptive Statistics


```{r}
library(tidyverse)
library(readxl)
library(writexl)
library(stringr)
library(gt)
```

# load clean exit survey data 2019-2023 (created in another script)
```{r}
sub_e2019_2023 <- readRDS("sub_e2019_2023.rds")
```

```{r}
setwd("C:/Grad School/Student Affairs/Exit Survey")
```

*Clean Data*

# create subset for data and investigate variables 
```{r}
debt <- sub_e2019_2023 %>% 
  select(debt_priorMPP, 
         debt_afterMPP, 
         year)

# 6 NA values 
sum(is.na(debt))

# 260 rows
summary(debt_clean)

# row # 's that contain NA's for debt_priorMPP: 24, 65, 162
which(is.na(debt$debt_priorMPP))

# row # 's that contain NA's for debt_afterMPP: 24, 65, 162
which(is.na(debt$debt_afterMPP))

# making sure that these values are both NA. They are all in year 2019
debt[c(24,65,162),] %>%  view()

# there are values for the rest of 2019
debt %>%  
  filter(year == 2019) %>%  
  view()

# remove the NA's from the data set 
debt_clean <- debt %>%  
  filter(!is.na(debt_priorMPP))

# There are 3 row numbers where the respondent said "Decline to state" for 
# both answers: 63, 194, 237. 
# There are two rows for debt_prior that says "Decline to state": 224, 232
# There is one row for debt_after that says "Decline to state": 36

sum(debt_clean$debt_priorMPP == "Decline to state")
which(debt_clean$debt_priorMPP == "Decline to state")

sum(debt_clean$debt_afterMPP == "Decline to state")
which(debt_clean$debt_afterMPP == "Decline to state")


# confirming that "Decline to state" is the value for rows: 36, 63. 194, 223, 232, 237
debt_clean[c(36, 63, 194, 
             224, 232, 237),] %>%  
  view()

# use the code above to remove these values 
debt_clean <- debt_clean %>%  
  filter(debt_priorMPP != "Decline to state",
         debt_afterMPP != "Decline to state")

# need to (finish?)
unique(debt_clean$debt_priorMPP)
unique(debt_clean$debt_afterMPP)
```

*Analyze Data*

# investigate debt_clean (count by year) 
```{r}
debt_clean %>%  
  group_by(debt_priorMPP) %>% 
  summarize(count = n()) %>%  
  arrange(debt_priorMPP)


debt_clean %>%  
  group_by(debt_afterMPP) %>% 
  summarize(count = n()) %>%  
  arrange(debt_afterMPP)

```

*Manipulate Data*

# create and clean debt_prior_count
```{r}
debt_prior_count <- debt_clean %>%  
  group_by(debt_priorMPP) %>%  
  summarize (count_prior = n())

# debt_priorMPP is missing two values, so creating observations to add to debt_prior_count
new_rows <- data.frame(
  debt_priorMPP = c(
    "$70,000 - $80,000", 
    "$80,000 - $90,000"), 
  count_prior = c(0, 0))

# add rows to debt_prior_count
debt_prior_count <- debt_prior_count %>%  
  rbind(new_rows) %>%  
  arrange(debt_priorMPP) 
```

# create and clean debt_after_count
```{r}
debt_after_count <- debt_clean %>%  
                      group_by(debt_afterMPP) %>%  
                      summarize (count_after = n())

# debt_afterMPP is missing one value
new_row <- data.frame(
                      debt_afterMPP = "$60,000 - $70,000", 
                      count_after = 0)

# add rows to debt_after_count
debt_after_count <- debt_after_count %>%  
  rbind(new_row) %>%  
  arrange(debt_afterMPP)

# both objects have the same number of rows 
nrow(debt_prior_count) 
nrow(debt_after_count)
```

# merge debt_after_count$count_prior to debt_after_count 
```{r}
combined_debt_count <- debt_prior_count %>%  
  mutate(count_after = debt_after_count$count_after)
```

# create object that groups values into 0, 1-9k, 10k-39k, 40k-79k, & 80k-120k
```{r}
grouped_barchart <- data.frame(
  debt_amount = c("0" , 
                  "1-9K",
                  "10k-39k", 
                  "40k-79k", 
                  "80k-120k"), 
  
  count_priorMPP = as.double(c(
                                  combined_debt_count$count_prior[10],
                                  combined_debt_count$count_prior[11], 
                               sum(combined_debt_count$count_prior[1:3]), 
                               sum(combined_debt_count$count_prior[4:7]),                                                  sum(combined_debt_count$count_prior[c(8:9, 12)])
                               )), 
  
  count_afterMPP = as.double(c( 
                                  combined_debt_count$count_after[10],
                                  combined_debt_count$count_after[11], 
                               sum(combined_debt_count$count_after[1:3]), 
                               sum(combined_debt_count$count_after[4:7]),                                                  sum(combined_debt_count$count_after[c(8:9, 12)])
                               ))) 

```

# restructure data to work with grouped barchart (create fill variable "prior/after")
```{r}
# gather functions transforms wide data into long format by stacking columns. 
format_grouped_barchart <- grouped_barchart %>%  
  gather(
          key = "before_after", 
          value = "Value",
          -debt_amount
          )

# recode variable names 
 clean_format_grouped_barchart <- format_grouped_barchart %>%  
   rename(debt_before_after = before_after, 
          count = Value)
 
# recode values for debt_prior_after to 'Before MPP' and "After MPP"
 clean_format_grouped_barchart <- clean_format_grouped_barchart %>%  
   mutate(debt_before_after = 
            recode(debt_before_after,
                   "count_priorMPP" = "Before MPP", 
                   "count_afterMPP" = "After MPP"
                   ))
 
 
```

# factor values for group_barchart 
```{r}
clean_format_grouped_barchart$debt_amount <-  
  factor(clean_format_grouped_barchart$debt_amount,
                             levels = c(
                               "0" ,
                               "1-9K",
                               "10k-39k", 
                               "40k-79k", 
                               "80k-120k"
                               ))

clean_format_grouped_barchart$debt_before_after <- factor(clean_format_grouped_barchart$debt_before_after, 
                              levels = c(
                                 "Before MPP", 
                                 "After MPP"
                                 ))
```

*Visualize Data* 

# create grouped barchart
```{r}
  ggplot(clean_format_grouped_barchart, 
          aes(x= debt_amount, 
              y = count, 
              fill = debt_before_after)) + 
  
  geom_bar(
        stat = "identity", 
        position = "dodge",  
        color = "black") +   
 
  scale_y_continuous(
        limits = c(0, 140),
        breaks = seq(0, 140, by = 20),
        labels = seq(0, 140, by = 20)) + 
  
  labs(
    x = "Amount of Debt",
    y = "Number of Students",
    title = "Student Debt Before and After MPP \n (2019-2023)",
    fill = "" ) +     
  
  scale_fill_manual(
    values = c("Before MPP" = "lightblue", 
               "After MPP" = "navyblue")) + 

  theme(
    axis.text = element_text(hjust = .5, size =10),
    plot.caption = element_text(hjust = 1 ),
    plot.title = element_text(hjust = .5, face = "bold", size = 12)) +
  
  labs(
    caption = "The only responses included are from the 260 students 
    who answered both questions.")

```

*Descriptive Statistics* 

# find percentage of students with 0 debt before/after program 
```{r}
# create an object that finds the percent of students that had a certain amount of debt 
percentage_count <- clean_combined_count %>% mutate(
  percent_prior = round(count_prior/sum(count_prior),2), 
  percent_after = round(count_after/sum(count_after),2))

# checking the total number of students 
sum(percentage_count$count_after)

# create subset for $40K to 80K 
forty_79 <- percentage_count %>%  filter(
  debt_priorMPP %in% c(
     "$40,000 - $50,000", 
     "$50,000 - $60,000",
     "$60,000 - $70,000", 
     "$70,000 - $80,000"
      ))                          

sum(forty_79$percent_prior)
sum(forty_79$percent_after)

mean(percentage_count$percent_prior)
summary(percentage_count$count_after)
```

# find rough estimate of average debt 
```{r}
average_count <- percentage_count %>%  mutate( 
  estimate = as.numeric(c(
    0, 
    5000,
    15000,
    25000,
    35000,
    45000, 
    55000, 
    65000, 
    75000, 
    85000, 
    95000, 
    110000)))

# create variable that multiplies count_after by the estimate. Then take mean to give rough estimate of total debt 
total_average_count <- average_count %>%  
  mutate(
    total_estimate = count_after*estimate)

# find mean (29019.23)
sum(total_average_count$total_estimate)/sum(total_average_count$count_after)
```

