---
title: "Stats_PS3"
author: "Sam Newman"
date: "2023-03-16"
output: pdf_document
editor_options: 
  markdown: 
    wrap: 90
---
Background: This script contains my answers for a graduate level statistics II course at UCLA. 
The course focused on various types of regression and methods to control for bias. 
The following types of statistical methods are used in the script: pooled regression,
fixed effects regression, removing outliers, multiple linear hypothesis testing, 
regressions with panel structured data, and testing for good controls.  

Outline: 

```{r, include = FALSE}
library(tidyverse)
library(lfe)
library(car)
library(haven) 
library(ivreg)
```

```{r, include = FALSE}
setwd("C:/Grad School/Year 1/Q2_Winter/Stats/Data")
```


```{r, results = "hide"}
current_pop_survey_78_95 <- read.csv("CPS78_85.csv")
murder_rates_state_87_90_93 <- read.csv("murder.csv")
law_salaries <- read.csv("lawsch85.csv")
rent_law <- read_dta("RENTAL_PP208.dta")
fertil2 <- read_dta("FERTIL2.DTA")
```

#investigate current_pop_survey_78_95 and murder_rates_state_87_90_93
```{r}
summary(current_pop_survey_78_95)
length(current_pop_survey_78_95)
nrow(current_pop_survey_78_95)

summary(murder_rates_state_87_90_93)
length(murder_rates_state_87_90_93)
nrow(murder_rates_state_87_90_93)

summary(law_salaries)
length(law_salaries)
nrow(law_salaries)

summary(rent_law)
length(rent_law)
nrow(rent_law)
```
# determine the effect education has on wage and control for other variables for the current_pop_survey_78_95
```{r}
lm1 <- lm(lwage ~ educ + female + married + union + exper +y85, current_pop_survey_78_95)

summary(lm1)
```

Interpretation: 

After controlling for sex, martial status, union affiliation, work experience, 
and year, each additional year of education increases wages, on average, by 
8.7568%. In the year 1985, individuals were earning 39.43% more than in 1978.  

# determine the effect of number of executions on murder rate and control for other variables(pooled regression)  
```{r}
lm2<- lm( mrdrte ~ exec + unem + d90 + d93, murder_rates_state_87_90_93)
summary(lm2)
```

Interpretation: 

After controlling for unemployment, an increase of 1 execution (over 3 years of 
data) increased a state's murder rate in the state by .1628, on average. 
 
After controlling for executions (within a state) and years 1987, 1990, and 
1993, a one percentage increase in the unemployment rate in a given year, 
increased in a state's murder rate by 1.3908 per 100,000 for the same year, on average. 

Since 19987 is our baseline, given that d87 is omitted from lm2, the year 1990
had 2.6733 (out of 100,000) more murders than 1987, and 1993 had 1.6073 (out
of 100,000) more more murders than 1987. Both interpretations are after
controlling for unemployment and executions in a state over a 3 years of data,
and represent an average of murder rates for the given year.

Explanation of Potential Bias:

There are other variables that are correlated with either executions or
unemployment (or potentially both) AND have a direct effect on murder rates.

A state's population has a positive correlation with the amount
of executions in a state. More people in jails increases the propensity for 
executions and more populous cities tend to have higher crime and murder rates. 
Therefore, the exclusion of population, in respect to executions, from the model 
would cause the effect of executions on murder rates to be underestimated and 
for there to be negative bias.

# regress number of executions onto murdates and account for fixed effects  
```{r}
lm3 <- felm (mrdrte ~ exec + unem + d90 + d93 | id , murder_rates_state_87_90_93)
summary(lm3)
```

Intrepretation: 

Fixed effects model accounts for unobserved, fixed characteristics that are
associated with individual states. For example, this 
model accounts for Alabama's population for the years 1987, 
1990, 1993. Alabama's population barley changed for these years (fixed) and 
wasn't explicitly controlled for in lm3 (unobserved), yet is accounted for 
in this model. 

After controlling for unemployment and years 1987, 1990, 1993, an increase of 
one execution (within a state) will decrease the murder rate by -.1383 
(out of #100,000) on average. We have removed bias from our pooled regression 
(lm2) because this model now accounts for unobserved, fixed characteristics for 
individual states (like Alabama's population) which have a correlation with
executions and a direct effect on murder rates. The change 
of Beta 1_Exec reflects the collective sum of these fixed, unobserved 
characteristics of individual states being added to our model.

# find the state with the most executions and 2nd most executions 
```{r}
#produce table for 1993 by state and executions
murder_rates_state_87_90_93 %>%  
  select(state, 
         exec, 
         d93) %>%  
  filter(d93 == 1) %>%  
  table()

#Confirm Texas had 34 executions in 1993
murder_rates_state_87_90_93 %>%  
  select(state, 
         exec, 
         d93) %>%  
  filter(d93 == 1, 
         exec >= 34)

#Confirm Virginia had 11 executions in 1993
murder_rates_state_87_90_93 %>%  
  select(state, 
         exec, 
         d93) %>%  
  filter(d93 == 1, 
         exec >= 11)
```

Answer: 

Texas had 25 more executions (34 total) than Virginia had (11) in 1993. 

# remove Texas from the data since it is an outlier and recalcuate fixed effects regression
```{r}
murder_noTX <- murder_rates_state_87_90_93 %>%  
  filter(state != "TX")

lm4 <- felm (mrdrte ~ exec + unem + d90 + d93 | id , murder_noTX)
summary(lm4)
```

Interpretation:

Executions as a deterrent for murder becomes more effective, although still
not statistically significant, as an increase in 1 execution (within a state) 
decreases murderer rate by .1897 (out of 100,000), on average after controlling
for unemployment. This is more than 0.1383 decrease in murder rate when Texas 
was included the model. This is because there is an outlier in 1993 when Texas 
had 34 executions, yet the murder rate for Texas that year was 11.9
(which is more than .6 decrease from their yearly average of 12.57 deaths
per 100,000). Thus, the dramatic increase in executions in 1993 did not lead 
to a proportional increase in murder rate, which decreases the effective rate 
of execution as a deterrent for murder. 


# test multiple linear hypothesis testing 
```{r}
lm5 <- lm(lsalary ~ LSAT + GPA + llibvol + lcost + rank, law_salaries)
summary(lm5)

linearHypothesis(lm5,c("LSAT = 0", "GPA = 0"))
```
Interpretation: 

We can be 99.9% confident that LSAT or GPA have an effect on future earnings 
(lsalary) given that our p-value (9.518e-05) is less than our alpha (.001). 


# determine the effect rent control has on rent and control for other variables 
```{r}
lm6 <- lm(lrent ~ rentcontrol + lpop + lavginc + pctstu + ltothsg + 
            rentcontrol*y90, rent_law)
summary(lm6)
```

Interpretation: 

Controlling for (log) city population, (log) average income, (percent) student 
housing, and (log) total housing units, when rent control was enacted 
(year 1990), rent control decreased rent by 8.4058%, on average. 

# determine the effect rent control has on rent while considering the panel structured date 
```{r}
lm7 <- felm(lrent ~ rentcontrol + lpop + lavginc + pctstu + ltothsg + y90 + rentcontrol*y90 | city, data = rent_law)

summary(lm7)
```

Interpretation: 

The coefficient for the interaction term of rent control and the year 1990, 
increased to -.113533. That is, rent control is now decreasing rent, in 1990, 
by 11.3533% on average, which is a bigger effect than when the city level 
panel was ignored (8.4058%). Since we are now accounting for the city level 
panel structure changes in individual cities from 1990 to 1980, aside from
rent control, are now being netted out, so more of the change in rent 
(dependent variable) is being explained by rent control.  

# determine if non-rent control cities are good controls for rent-controlled variables 
```{r}
lm8  <- lm(lpop    ~ rentcontrol + y90 + rentcontrol*y90, rent_law)
lm9  <- lm(lavginc ~ rentcontrol + y90 + rentcontrol*y90, rent_law)
lm10 <- lm(pctstu  ~ rentcontrol + y90 + rentcontrol*y90, rent_law)
lm11 <- lm(ltothsg ~ rentcontrol + y90 + rentcontrol*y90, rent_law)

summary(lm8)
summary(lm9)
summary(lm10)
summary(lm11)
```

Interpretation: 

Non-rent control cities are good controls because the coefficient of the 
interaction term of rent control and year 90 is close to 0 (not statistically significant). 
For example in 1990, a cities' rent control law only decreased it's population 
by .13584%, but that number has a large standard error and a p-value of .996. 
Therefore, we would fail to reject H0 (Beta3_Rent Control*1990 = 0), and 
conclude that rent control didn't effect the population of a city, and hence population of a city 
is good control for cities that do and do not receive rent control, in years 1980 and 1990. 