---
title: "PP_prepared ranking"
author: "Sam Newman"
date: "2023-09-05"
output: pdf_document
---

Background:
The data comes from one quantitative question asked in a survey that was given to
a universities (named disclosed for confidentiality) 2nd year MPP students (n = 269) 
from 2019-2023. The question asked students to rank how prepared they feel for their
career after the graduate program from "1= Not Prepared" to "5 = Very Prepared". 
This script uses a for loop to calculate the weighted mean for each year's score (2019-2023). 
Then each year's distance from the mean, in terms of standard deviation, is 
calculated to determine the statistical significance of the result. 
A chart is then created in Excel displaying the information.  

Outline: 
1. Descriptive Statistics
    -Find weighted mean for all years 
    -Use for loop to find weighted mean for each year 

2. Manipulate Data 

3. Descriptive Statistics
    -Find each years distance from the weighted mean in terms of standard deviation 


```{r}
library(tidyverse)
library(readxl)
library(writexl)
library(stringr)
library(gt)
```

```{r}
setwd("C:/Grad School/Student Affairs/Exit Survey")
```

# load clean exit survey data 2019-2023 (created in another script)
```{r}
sub_e2019_2023 <- readRDS("sub_e2019_2023.rds")
```

# create subset for data and investigate  
```{r}
#create object called ranking. w/ variables of interest
pp_ranking <- sub_e2019_2023 %>%  
  select(pp_prepared_ranking, 
         debt_afterMPP, 
         year)

#there are 0 NAs
sum(is.na(pp_ranking$pp_prepared_ranking))

#PP_ranking ranges from 1-5. 269 observations 
summary(pp_ranking)
```

*Descriptive Statistics*

# calculate the weighted mean for all years 
```{r}
#count values in ranking across ALL years. 269 observations   
count_pp_ranking_all_years <- pp_ranking %>%  
  group_by(pp_prepared_ranking) %>%  
  summarize(count = n()) 
 
w_mean_pp_ranking_all_years <- count_pp_ranking_all_years %>%  
  mutate(weighted_rank = 
           sum(pp_prepared_ranking * count)/sum(count)
  )
```

# use for loop to calculate weighted mean years 2019:2023
```{r}
range <- c(2019:2023)

count_pp_ranking <- list()
w_mean_pp_ranking <- list()

for (i in range) {
  count_pp_ranking[[as.character(i)]] <- pp_ranking %>% 
    filter(year == i) %>%  
    group_by(pp_prepared_ranking) %>%  
    summarize(count = n())
  
  w_mean_pp_ranking[[as.character(i)]] <- count_pp_ranking[[as.character(i)]] %>%  
    mutate(weighted_rank = sum(pp_prepared_ranking * count) / sum(count))
}
```

*Manipulate Data* 


# unnest list 
```{r}
w_mean_pp_ranking_2019 <- w_mean_pp_ranking[[1]]
w_mean_pp_ranking_2020 <- w_mean_pp_ranking[[2]]
w_mean_pp_ranking_2021 <- w_mean_pp_ranking[[3]]
w_mean_pp_ranking_2022 <- w_mean_pp_ranking[[4]]
w_mean_pp_ranking_2023 <- w_mean_pp_ranking[[5]]
```

# create object containing average for years 2019:2023 and the average of 2019:2023
```{r}
  cumulative_w_mean_pp_ranking <- as.data.frame(
    c(
              w_mean_pp_ranking_all_years$weighted_rank[1], 
              w_mean_pp_ranking_2019$weighted_rank[1], 
              w_mean_pp_ranking_2020$weighted_rank[1], 
              w_mean_pp_ranking_2021$weighted_rank[1], 
              w_mean_pp_ranking_2022$weighted_rank[1], 
              w_mean_pp_ranking_2023$weighted_rank[1])
              )

#change column name
clean_cumulative_w_mean_pp_ranking <- cumulative_w_mean_pp_ranking %>%  
  rename(
   "weighted_mean" = "c(w_mean_pp_ranking_all_years$weighted_rank[1], w_mean_pp_ranking_2019$weighted_rank[1], w_mean_pp_ranking_2020$weighted_rank[1], w_mean_pp_ranking_2021$weighted_rank[1], w_mean_pp_ranking_2022$weighted_rank[1], w_mean_pp_ranking_2023$weighted_rank[1])") %>%  
  mutate(
    years = c(
      "All Years", 
      2019:2023), 
    weighted_mean = 
      round(weighted_mean, 2)) %>%  
  select(
    years, 
    weighted_mean)
 
```

*Descriptive Statistics* 

# determine level of statistical significance using standard deviations 
```{r}
#This uses the weighted means for each year 
sd(clean_cumulative_w_mean_pp_ranking$weighted_mean)

## of standard deviations away from the mean (add to object)
standard_deviation <- data.frame( 
  Years = c(
      "All Years", 
      2019:2023), 
  "Standard Deviations from Average" = 
    round(abs(clean_cumulative_w_mean_pp_ranking$weighted_mean[1]-      clean_cumulative_w_mean_pp_ranking$weighted_mean)/
    sd(clean_cumulative_w_mean_pp_ranking$weighted_mean),
  2))


```

