---
title: "Qualitative_CA Pop Race"
author: "Sam Newman"
date: "2023-06-27"
output: html_document
---

Background: 
This script was part of an assignment to update California's population count from the U.S. Census Bureau (https://data.census.gov/table/ACSDP5Y2022.DP05?g=860XX00US06902). The data is cleaned, manipulated, and presented in a pie chart. The data is subseted to include seven prominent races in California. 

Outline: 

1. Clean Data
2. Descriptive Statistics
3. Visualize Data 

```{r}
library(tidyverse)
library(readxl)
library(ggrepel) #helps with placing labels outside of pie chart 
library(ggtext)  #helps align caption for ggplots
```

```{r}
setwd("C:/Grad School/The Campaign/Candidate Skills Assessment/data")
```
```{r}
CA_pop_2020 <- read_xlsx("CA_pop_2020.xlsx", 
                         sheet = "Data", 
                         col_names = c(
                           "race", 
                           "population", 
                           "margin of error",
                           "percent", 
                           "percent margin of error")
                         ) 
```

*Clean Variable*  

# subset and clean variable 
```{r}
subset_CA_pop_2020 <- CA_pop_2020 %>%  
  select(race, population) %>% 
  slice(c(77,83:89),) %>%  #races of interest
  mutate(population = as.numeric( 
           gsub(",", "", population))) # remove commas

summary(subset_CA_pop_2020)
nrow(subset_CA_pop_2020)
length(subset_CA_pop_2020)
```



# add 'Native Hawaiian and Other Pacific Islander' to 'Asian' 
```{r}
#create df with just population for Asian and Native Hawaiian and Other Pacific Islander" 
AANHPI <- subset_CA_pop_2020 %>% 
  slice(5:6) 

#add new row summing the population and creating a combined name for race 
totaled_AANHPI<- AANHPI %>%  
  add_row( 
    race = "AANHPI", 
    population = sum(AANHPI$population))

#only keep total
only_totaled_AANHPI <- totaled_AANHPI[3,]

#add total back to subset_CA_pop_2020
subset_CA_pop_2020 <- rbind(
  subset_CA_pop_2020, 
  only_totaled_AANHPI)

#remove rows
subset_CA_pop_2020 <- subset_CA_pop_2020 %>%  
  filter(race != c(
    "Asian alone", 
    "Native Hawaiian and Other Pacific Islander alone")
         )
```

# Change names of race values 
```{r}
clean_subset_CA_pop_2020 <- subset_CA_pop_2020 %>%  
  mutate(race = recode(race,
                       "White alone" = "White", 
                       "Black or African American alone" = "Black", 
                       "American Indian and Alaska Native alone" = "AIAN", 
                       "Hispanic or Latino (of any race)" = "LatinX",
                       "Two or more races" = "Multiracial", 
                       "Some other race alone" = "Other"))
```
*Desriptive Statistics* 

# create percent column 
```{r}
percent_subset_CA_pop_2020 <- clean_subset_CA_pop_2020 %>%  
  mutate(percent = 
           population/sum(population)) %>%
  arrange(desc(population))
```

*Visualize Data* 

# create vector to store colors for piechart 
```{r}
colors <- c("forestgreen",
            "orange",
            "cyan",
            "gray", 
            "blue",
            "yellow",
            "purple")
```


# reformat data to include positions column (for labels)
```{r}
pie_pop_2020 <- percent_subset_CA_pop_2020 %>% 
  mutate(csum = rev(cumsum(rev(population))), 
         pos = population/2 + lead(csum, 1),
         pos = if_else(is.na(pos), population/2, pos))
```

# make piechart 
```{r}
ggplot(percent_subset_CA_pop_2020, 
       aes(x = "", 
           y = population, 
           fill = fct_inorder(race))) +
  geom_col(width = 1, 
           color = "white") +
  coord_polar(theta = "y") +
   scale_fill_manual(values = colors) +
  geom_label_repel(data = pie_pop_2020,
                   aes(y = pos, 
                       label = 
                         paste0(race, "\n", " ", signif(percent*100, digits = 2), "%")),
                   size = 2.5, 
                   nudge_x = 1) +
  theme_void() + 
  theme(legend.position = 'none', 
        plot.title = element_text(hjust = 0.5, 
                                  size = 11, 
                                  face = c("bold.italic")),
        plot.caption = element_text(hjust = .5)) + 
  ggtitle("California Population by Race") + 
  labs(caption = 
         ~bold("Source:")~"U.S. Census Bureau, American Community Survey, 2020 5-year Estimates, DP05.") 
```
