---
title: "Quantitative_HS Grad Rates"
author: "Sam Newman"
date: "2023-06-27"
output: html_document
---

Background: 
This script was part of an assignment to update California's high school graduation rates from the California Department of Education (https://www.cde.ca.gov/ds/ad/datasummary.asp). The data is cleaned, and visualized in a horizontal bar chart. 

Outline: 

1. Clean Data
2. Visualize Data 

```{r}
library(tidyverse)
library(readxl)
```

```{r}
setwd("C:/Grad School/The Campaign/Candidate Skills Assessment/data")
```

```{r}
full_hsgrad_2020_2021 <- read_xlsx("CA_HS_grad_2021-2022.xlsx")
```
*Clean Data* 

# subset and clean data 
```{r}
subset_hsgrad <- full_hsgrad_2020_2021 %>%  
  select("Race / Ethnicity", 
         "Cohort Students", 
         "Regular HS Diploma Graduates", 
         "Cohort Graduation Rate") %>%  
  
  rename(
    "Race" = "Race / Ethnicity", 
    "TotalStudents" = "Cohort Students",
    "GraduatingStudents" = "Regular HS Diploma Graduates",
    "GraduationRate" = "Cohort Graduation Rate") %>%  
  
  mutate(
    TotalStudents = 
      as.numeric(TotalStudents), 
    GraduatingStudents = 
      as.numeric(GraduatingStudents), 
    GraduationRate = 
      as.numeric(GraduationRate))
```


# combine 'filipino', 'pacific islander', and 'asian' to AANHPI
```{r}
#subset Filipino, pacific islander, and asian 
AANHPI <- subset_hsgrad %>% 
  slice(3:4,6)

totaled_AANHPI <- AANHPI %>%  
  add_row(
    Race = "AANHPI", 
    TotalStudents = 
      sum(AANHPI$TotalStudents), 
    GraduatingStudents= 
      sum(AANHPI$GraduatingStudents), 
    GraduationRate = 
      mean(AANHPI$GraduationRate))

only_totaled_AANHPI <- totaled_AANHPI[4,]

#add total back to clean_subset_hsgrad
clean_subset_hsgrad <- rbind(
  subset_hsgrad, 
  only_totaled_AANHPI)

#remove Filipino, Pacific Islander, Asian, and Statewide Total 
clean_subset_hsgrad2 <- clean_subset_hsgrad[c(-3,-4,-6, -10),]
```

# rename race values and add commas
```{r}
clean_subset_hsgrad2 <- clean_subset_hsgrad2 %>%  
  mutate(
    Race = recode(Race, #rename values
                  "African American" = "Black",
                  "American Indian or Alaska Native" = "AIAN",
                  "Hispanic or Latino" = "Latinx"), 
   
   TotalStudents = #add commas
     prettyNum(TotalStudents,
               big.mark = ",", 
               scientific = FALSE))
```


*Visualize Data* 

# order bars via factoring to match graph 
```{r}
factor_subset_hsgrad2 <- clean_subset_hsgrad2

factor_subset_hsgrad2$Race <- factor(
  factor_subset_hsgrad2$Race, 
  levels = c(
    "Not Reported", 
    "White",
    "Two or More Races",
    "Latinx",
    "Black", 
    "AIAN", 
    "AANHPI"))

```

# create barchart 
```{r}
ggplot(factor_subset_hsgrad2,
       aes(x = Race, 
           y = GraduationRate, 
           fill = Race, 
           label = paste0(TotalStudents,
                          "    ", 
                          scales::percent(round(GraduationRate,2))))) +
  coord_flip() +
  geom_col(position = 'dodge', 
           width =.7) +
  geom_text(position = 
              position_dodge(width =.9),
              hjust = .6, 
              size = 3.5, 
              fontface = "bold") + 
  scale_fill_manual(values  = c("red",
                                "orange", 
                                "yellow", 
                                "green",
                                "lightblue",
                                "grey",
                                "purple")) +
  scale_y_continuous(
    labels = scales::percent, 
    limits = c(0,1), #sets tick mark range
    breaks = seq(0,1,.1)) +
  
  theme_classic() +
  theme(
    axis.title.x = element_blank(),
    axis.title.y = element_blank(),
    legend.position = "none") + 
  
  ggtitle("High School Graduation Rates by Race/Ethncity") + 
  theme(
    plot.title = element_text(
      hjust =.5, 
      size = 11, 
      face = c("bold.italic")), 
    plot.caption = element_text(hjust = 0)) +
  labs(caption ="\n" ~bold("Source:")~"California Department of Education, 2020-2021 Four-year Adjusted Cohort Graduation Rate.")
```
